# The Chi-square Distribution {#chisquare}


```{r}
library(tidyverse)
```
## Background

The chi-square permeates biostatistics in several important ways.  

For example, the $\chi^2$ distribution is used to generate p-values for Pearson's chi-square test statistic, \[\chi^2=\sum_{i=1}^{n}\frac{(O_i-E_i)^2}{E_i}\] which is used in goodness-of-fit and independence tests.

The $\chi^2$ distribution is used to generate p-values for tests of homogeneity and also to calculate the confidence intervals of standard deviations.

A good way to think of the chi-square distribution more generally is as a probability model for the sums of squared variables. As such, the $\chi^2$ test statistic only takes on positive values.

If \[X_1,..,X_k\] is a set of standard normal variables, then the sum of their squared values is a positive, random variable $Q$, \[Q=\sum_{i=1}^{k}{X^2_i}\] that will take on a $\chi^2$ distribution with $k$ degrees of freedom.

The mean of any $\chi^2$ distribution is $k$ while it's variance is $2k$. The distributions are skewed (mode = $k-2$) except that those having very large degrees of freedom, which are approximately normal.

## dchisq

`dchisq` is the $\chi^2$ probability density function in R. 

The simplest way to think of `dchisq` is as the function that gives you the probability distribution of the $\chi^2$ test statistic.

The $\chi^2$ probability density function is:


$$p(x)=\frac{e^\frac{-x}{2}x^{\frac{k}{2}-1}}{2^\frac{k}{2}\Gamma(\frac{k}{2})}$$
Given  a chi-square value and the degrees of freedom of the dataset as input, `dchisq` returns the probability for a given chi-square value.  

For example, if a 2X2 test of independence (df=1) returns a $\chi^2$ value of 4, the probability of obtaining that value (its "density") is equal to 0.02699:

```{r}
dchisq(x=4, df=1)
```

For continuous distributions like the $\chi^2$ such point density values are usually not particularly useful. In contrast, inspecting how the function behaves over a range of $\chi^2$ values and at different df's is informative.

That's plotted below:

```{r}
df = 9
x = seq(1,16,0.05)
pxd <- matrix(ncol=df, nrow=length(x))

for(i in 1:df){
  pxd[,i] <- dchisq(x, i)
}

pxd <- data.frame(pxd)
colnames(pxd) <- c(1:df)
pxd <- cbind(x, pxd)
pxd <- gather(pxd, df, px, -x)

ggplot(pxd, aes(x, px, color=df)) +
  geom_line() + xlab("chi-square")+ylab("p(chi-square)")
```

## pchisq

`pchisq` is the $\chi^2$ cumulative distribution function in R. 

The simplest way to think about the `pchisq` function is as the probabilities under the $\chi^2$ curve. 

The $\chi^2$ cumulative distribution function is:

$$p(x)=\frac{\gamma(\frac{k}{2}\times\frac{x}{2})}{\Gamma(\frac{k}{2})}$$
It returns the cumulative probability for an area under the curve up to a given $\chi^2$ value.

For example, the probability that a $\chi^2$ value with 1 degree of freedom is less than 4 is 0.9544997:

```{r}
pchisq(q=4, df=1, lower.tail = T)
```

```{r}
df = 9
x = seq(1,16,0.05)
pxd <- matrix(ncol=df, nrow=length(x))

for(i in 1:df){
  pxd[,i] <- pchisq(x, i, lower.tail = T)
}

pxd <- data.frame(pxd)
colnames(pxd) <- c(1:df)
pxd <- cbind(x, pxd)
pxd <- gather(pxd, df, px, -x)

ggplot(pxd, aes(x, px, color=df)) +
  geom_line() + xlab("chi-square")+ylab("cumulative p(chi-square)")
```


### Calculating p-values from pchisq

The `pchisq` function is used to calculate a p-value.

A p-value is a probability that a $\chi^2$ value, x, is as large or larger. \[p-value = P(\chi^2 \ge x)\]

This can be calculated using the `pchisq` function simply by changing lower.tail argument in the function to`lower.tail = F`

```{r}
pchisq(q=4, df=1, lower.tail = F)
```
Here p-values for the $\chi^2$ at various df's: 

```{r}
df = 9
x = seq(1,16,0.05)
pxd <- matrix(ncol=df, nrow=length(x))

for(i in 1:df){
  pxd[,i] <- pchisq(x, i, lower.tail = F)
}

pxd <- data.frame(pxd)
colnames(pxd) <- c(1:df)
pxd <- cbind(x, pxd)
pxd <- gather(pxd, df, px, -x)

ggplot(pxd, aes(x, px, color=df)) +
  geom_line() + xlab("chi-square")+ylab("p-value")
```

## qchisq

`qchisq` is the inverse of the $\chi^2$ cumulative distribution function.

This function takes a probability value as an argument, along with degrees of feedom, and returns a $\chi^2$ value corresponding to that probability.

Here's the $\chi^2$ value corresponding to the 95th percentile of a distribution with 3 degrees of freedom:

```{r}
qchisq(0.95, 3)
```

Here is what the quantile $\chi^2$ distribution looks like for several different df's:

```{r}
df = 9
p = seq(0.01, .99, 0.01)
xpd <- matrix(ncol=df, nrow=length(p))

for(i in 1:df){
  xpd[,i] <- qchisq(p, i)
}

xpd <- data.frame(xpd)
colnames(xpd) <- c(1:df)
xpd <- cbind(p, xpd)
xpd <- gather(xpd, df, qchisq, -p)

ggplot(xpd, aes(p, qchisq, color=df)) +
  geom_line() + xlab("probability")+ylab("chi-square")
```

## rchisq

R's `rchisq` function is used to generate random values corresponding to $\chi^2$-distributed data. This has utility in simulating data sets with skewed values, for example, to mimic overdispersed Poisson distributions.   

Here are 10 random values from a $\chi^2$ distribution with 3 degrees of freedom: 

```{r}
rchisq(10, 3)
```

