# Two-way ANOVA RM/CR {#twowaymixed}

```{r echo=TRUE, message=FALSE, warning=FALSE}
#library(magrittr)
library(tidyverse)
library(DescTools)
library(ez)
library(nlme)

```
Two-way ANOVA experiments allow for simultaneously testing the effects of two predictor variables. 

In a two way completely randomized ANOVA, all of the levels of both factors are randomly allocated to individual experimental units. Each experimental unit represents an independent replicate and each generates a single measurement for analysis.

In a two-way related/repeated measures ANOVA, all of the levels for both factors are measured within a single experimental unit. The experimental unit represents the independent replicate, and therefore the data set contains many intrinsically-related measurements from each replicate.  

A hybrid of the two above is called a two-way mixed ANOVA experimental design. In a mixed ANOVA, one factor is completely randomized, whereas the other is structured as a related/repeated measures within each replicate.

The ChickWeight study is a good example for visualizing how a two-way mixed ANOVA would operate.

## ChickWeight Dataset

The ChickWeight dataset in base R is from a study testing the effect of 4 diets on the growth of chicks. You can learn more about it by typing `?ChickWeight` in the console.

As you might imagine, the principal goal of a study like this is to determine if any of the 4 diets differ from each other in terms of how they impact chick weight over time.

This is a two way mixed ANOVA design. The two factors are diet and time. Chicks are randomly assigned to a diet. While on the diet each chick is weighed repeatedly at various time points. 

Therefore, Diet is a completely randomized (CR) factor, whereas the factor time (in units of days) is a related/repeated measure variable.

If that's not entirely clear, it should become more obvious by looking at a plot of the data.

## Munge ChickWeight data

We have several things to do before the ANOVA can be run. First, let's just look at the data graphically. That makes the study design a bit easier to understand.

Then we'll clean it up for what we need to do.

Step 1: Plot it for a quick visualization

Note how each point-to-point line represents a unique chick replicate. You can get a sense from this visual how weight correlates with the replicates by time. Heavier chicks are heavy throughout the time course, lighter chicks are lighter throughout the time course. This serves as an interesting visualization of biological variation.

Visually, can you tell if any of the diets differ from the others?

```{r}
ggplot(ChickWeight, 
       aes(Time, 
           weight,
           group=Chick, 
           color=Chick)
       ) +
  geom_point(
    
  ) + 
  geom_line(
    
  ) + 
  scale_color_viridis_d(
    
  ) +
  facet_grid(
    ~Diet
    ) 
```

Step 2: First, check for missing data on the within factor, chicks. 

We can't have any for a RM design. ezDesign is a slick function in the ez package with which to check for missing values:

```{r}
ezDesign(
    data = ChickWeight,
    x = .(Time),
    y = .(Chick), 
    row = NULL, 
    col = NULL, 
    cell_border_size = 1
)
```
Step 3: Remove Chicks for which full Time series is missing, while creating a repaired dataset. 

Here's a hemi dplyr-based way to do that:

```{r}
temp <- as.data.frame(
  ChickWeight%>%
  group_by(Chick)%>%
  tally()
)
# show only the Chick replicates that have incomplete measures on time
temp[temp$n<12, ]

# create new dataframe for ANOVA analysis in which the incomplete replicates are removed

ChickWeight1 <-  ChickWeight[!(ChickWeight$Chick %in% temp$Chick[temp$n<12]),]

```

Step 4: Check to make sure the new dataframe from the previous step is clean, that incomplete replicates are out

```{r}
ezDesign(
    data = ChickWeight1
    , x = .(Time)
    , y = .(Chick)
    , row = NULL
    , col = NULL
    , cell_border_size = 1
)
```

Step 5: The ChickWeight dataframe has Time as a numeric, which needs to be converted to a factor to get stuff to work.

```{r}
ChickWeight1$Time <- as.factor(ChickWeight1$Time)
```

One last check of the data structure:

```{r}
head(ChickWeight1)
str(ChickWeight1)
```

Done with the munge!

## The test

The main scientific objective of this study is to learn whether growth differs between the four diets. 

The study design calls for a two-way mixed ANOVA analysis as an omnibus test. If that passes, we'll have access to compare the diets post-hoc.

We're not particularly interested in the effect of the Time factor, per se. We know the chicks will grow with time. Nor are we particularly interested in the interaction effect between Time and Diet. Which is to say that we don't want to know at what times the various levels of the diets differ.

We just want to know if the diets differ.

Having said that, the repeated measures of weight on time seems, intuitively, a solid way for designing the experiment, compared to a few conceivable alternatives. 

It collects 12 different weights per replicate than a study that only weighed each critter once, say, on the 21st day. Thus, there is more thorough information on how each replicate responded to the diet. Which seems useful. 

But we're not explicitly applying a rate model to the data to derive growth rate constants that might be compared. The growth rate is more implicit...it's the information buried within the repeated measures. 

And we're also not particularly interested in knowing at what specific time points the various diets might differ.  

We just want to know if the four diets differ from each other.

Here's how that's done.

To encode a two-way mixed ANOVA using ezANOVA, simply configure the arguments such that the repeated measure factor is `within` and the completely randomized factor is `between`. An object named `two_wayMM`, which could have been named foo, is used to store the ANOVA output.

```{r message=FALSE, warning=FALSE, paged.print=FALSE}

two_wayMM <- ezANOVA(data = ChickWeight1,
                     dv = weight,
                     wid = Chick,
                     between = Diet,
                     within = Time,
                     type = 3,
                     detailed = T
                     )
two_wayMM
```
 
## Interpreting the ANOVA output

 Here ezANOVA creates Four objects:
 * $ANOVA
 * $Mauchly's Test for Sphericity
 * $Sphericity Corrections`
 * $aov (not shown)

### $ANOVA: The ANOVA table

Before partioning for sujects, the total residual sums of squares in the sample would be the sum of the residual error terms, or $295322.5 + 313495.0 = 608817.5$. 

Because weight measurements were repeated on each chick over time, this procedure allows us to account for some of the variation in the Time variable as that due to variation among the chicks. 

Pulling 313495 out from the overall residual provides a new (and lower than otherwise) residual term for the effect of Diet, while lowering the residual term for the effect of Time and the interaction.

The plot above gives you a sense of what's going on. What's really noticable, I think, is a lot of chick-to-chick variation, irrespective of the diet. There's also a lot of variation across the time frame. But the chick-to-chick variation grows really prominent near the end of the time course. 

What this does is help to isolate from the variation over the time course that which is due to the differences between the 45 chicks. It doesn't explain why that variation between the chicks occurs, its obviously biological, but it does allow us to partition it out from the Time effect.  

In the ANOVA table the intercept effect is not particularly important except for the SSd, which tabulates the SS that have been partitioned to the chicks. It is the subject effect. You can think of it as the average subject effect. An extreme value means that there is substantial variation within the chick replicates. Of course there is, they're growing up!

The second effect is Diet, which is important for all the reasons above. There are four diets and thus 3 degrees of freedom for the F test numerator. There are 45 chicks serving as replicates spread out into the four diet groups. The four diet group means cost a total of 4 degrees of freedom, leaving 41 degrees of freedom for the Diet residual variance (denominator). F has been calculated as \[F=\frac{\frac{SSn}{df_n}}{\frac{SSd}{df_d}}\]

The p-value is the probability of getting an F value as or more extreme from the null F(3,41) distribution. Given the p-value is less than a type 1 error threshold of 5%, we can reject that the F test statistic obtained in the sample belongs to the null. 

Normally we would conclude that this means the diets effect on weight differs. But that's hard to call to make given the interaction effect is also positive.

The general eta-squared (ges) value indicates that 16% of the overall variation is attributable to the Diet effect. 

The second effect is that for the twelve levels of the variable Time. The interpretation of it's table is much the same. There are seven time points per replicate, and one is lost to calculate the grand time mean, leaving 11 df for this test. The residual $df_d=(45-1)*(12-1)-(12-1)*(4-1)=451$. Almost 77% of the total variation in the sample can be explained the effect of time. Although a large effect, it isn't very interesting scientifically. There's no surprise that the chicks grew a lot!

The third effect is the interaction between the Diet and Time variables. $df_n=(4-1)*(12-1)=33$. The interaction between these two variables explains about 11.8% of the total variation in the sample. This suggests the various diets have some differential influence on the growth rates. It isn't just the diets, and it isn't just the time; the two factors together combine to change weights in a way that neither does alone.

### Mauchly's Test and Corrections

You'll recall one of the assumptions to be met for valid ANOVA analysis is that the response variances of the different levels of a factor should be about equivalent (homoskedasticity). This concept extends to the related measures scenarios. 

Sphericity is defined as uniform variance for the **differences** between factor levels. Sphericity is an assumption of RM ANOVA, and if violated, should be corrected. 

Quick inspection of the time course shows violation. Focus on the connecting lines. They have far greater slope near the end of the time course compared to early stages.

Mauchly's test, with its extreme F-statistic value, indicates we can reject the null that sphericity is present. There are differences in variance of differences across the time scale. The Df's for this test are Dfn*GGe & Dfd*GGe.

Because of that, we correct the Time result for sphericity using the Geiser-Greenhouse correction. We do this by using the value of p[GG] as a p-value for our test. The correction is dramatic, at least with respect to the p-value correction for the interaction effect. The p[GG]= 0.01045 for the interaction test is that which should be used for inference.

## Post hoc pairwise tests

A positive F-test opens the gate to perform post hoc analysis of any groups that interest us. The F tests for the two factors and the interaction shows that there are effects of time and diet on the growth of chicks. The effect of time is not particularly interesting since it is no surprise that chicks grow.

What the F tests haven't told us is which diets differ. Is one better than any other? Is one worse than any other?

The approach that we'll use to answer these questions is the TukeyHSD test, which is used to make a variety of comparisons between groups.  The only groups that are of interest scientifically, however, are those associated with the main effects of the diet. Even so, we should make those comparisons in the context of the overall experimental design.

The TukeyHSD test runs in three steps.

The first step is to create a linear model of the design. The second step is to create an anova object from that model. And the third step is the reason for the first two. The TukeyHSD test function takes an aov object as its data source before running the group comparisons.

Because it has a related measures component, the two-way mixed ANOVA experimental design is also known as a general linear mixed-effects model. Exactly what this means will be more clear when we get to discussing general linear models.

We use the `lme` function from the `nlme` package to regress the model on the ChickWeight1 dataset, generating a large table of coefficients for all the model elements. 

```{r}
model <- lme(weight ~ Diet + Time + Diet*Time, data=ChickWeight1, random = ~ 1|Chick)
model

```

Next, that linear model is passed into the `aov` function, to create an aov object. Why not use ezANOVA to create this? It doesn't create an object of the proper R class for the posthoc function. This is passed

Finally, we run the TukeyHSD test. This function is in the `DescTools` package. We specify Diet so that it only returns all ofthe possible comparisons between Diet groups. (If you remove the list argument it will return all posssible combinations of every group in the experiment! ) 

```{r}
bark <- lme(weight ~ Diet + Time + Diet*Time, data=ChickWeight1, random = ~ 1|Chick)
```


```{r}
PostHocTest(aov(bark,data=ChickWeight1), which="Diet", method="hsd")
```

What I like about the TukeyHSD test run this way is that it yields effect size, confidence intervals, and adjusted p-values for each comparison. For example, the difference in mean weight between the first and the second diets is 14.9 grams (95%CI(adjusted): 4.4 to 25.5 grams, adjusted p-value=0.0016).

### Here's what's been discovered

Thus, in terms of how the 4 levels of the Diet factor differ in causing chick weight we can conclude: a) all diets are better than Diet 1, b) Diets 3 and 4 are better than Diet 2, c) Diets 3 and 4 do not differ. 

**Reporting** 

"The data were analyzed by type 3 two-way mixed ANOVA with repeated measures on Time (R 3.5.2, ezANOVA). The diet, time interaction F test was positive following correction for a positive Mauchly's test (correction factor = 0.114145, GG-adjusted p-value = 0.01045). Post hoc comparisons using TukeyHSD adjustments for multiple comparisons indicate the mass of animals on Diet 1 is less than on all others, and that weight is greater on Diets 3 and 4 than on Diet 2." 

Ideally, report the last sentence in a table, showing the results. 

*It's very important to understand here that we took the p-value of p[GG] from the sphericity correction table, rather than that for Time from the ANOVA table, as our inference for this experiment*

